---
title: "snp_analysis_nv_pa"
output: github_document

---
Once you have a VCF you can read it in and make it a genlight object.  Follow that with making a PCA.  You will need meta data for each sample (sample name and population)

```{r setup, include=FALSE}
library(tidyverse)
library(vcfR)
library(ggplot2)
library(adegenet)
library(poppr)
library(RColorBrewer)
library(ggrepel)
```

read in VCF and match to metadata table.

```{r}

#read in vcf calcualated using freebayes
Bd.VCF <- read.vcfR("NV_PA_Bd_wrefs_freebayes_trimmed_filtered.vcf")

#read in file with sample metadata
read_csv(file = "Bd_PA_NV_meta_trim_final.csv") -> Bd.meta

#join in a meta table based on sample ID
colnames(Bd.VCF@gt)[-1] -> vcf.names
as.data.frame(vcf.names) -> vcf.names
colnames(vcf.names) <- "Sample_ID"
left_join(vcf.names, Bd.meta, by = "Sample_ID") -> vcf.meta

#check
all(colnames(Bd.VCF@gt)[-1] == vcf.meta$Sample)

```


Make VCF object a genlight object.  Set populations.  Make PCA.  use the function subpop() to subset you factor levels if you want!

```{r}
gl.Bd <- vcfR2genlight(Bd.VCF)
ploidy(gl.Bd) <- 2
pop(gl.Bd) <- vcf.meta$GROUP

#get summary of data
gl.Bd

```

now to do a DAPC using adegenet

```{r}
#finds clusters in data
grp <- find.clusters(gl.Bd, max.n.clust=10, n.pca = 100, choose.n.clust = F, criterion = "diffNgroup")

#run with 50 PCs 
dapc1 <- dapc(gl.Bd, grp$grp, n.pca=50, n.da=1)
summary(dapc1)
scatter(dapc1)

#now this may be overfitting because we are using so many PCs. So lets find out how many PCs we should use

#then use this to find the optimal number of PCs to use
temp <- optim.a.score(dapc1)

#ran this 10x and got the following optimal alpha scores:
#1, 1, 1, 1, 1, 1, 1, 1, 1, 1, 1

#run the DAPC again now we will just use 2 PCs
dapc1 <- dapc(gl.Bd, grp$grp, n.pca=2, n.da=1)

#to explore split
scatter(dapc1)

#to get summary
summary(dapc1)
#to see assignment probabilities
assignplot(dapc1)

#For instance,which are the most ’admixed’ individuals? Let us consider as admixed individuals having no more than 99% of probability of membership in a single cluster:
 
unassign <- which(apply(dapc1$posterior,1, function(e) all(e<0.99)))
unassign <- as.numeric(unassign)

assign1 <- as_tibble(as.numeric(dapc1$assign))
assign1[unassign,] <- 0

#sometimes the 1/2 values are switched and don't necessarily correspond to GPL1/2. Go back and check and fix them with this code if needed.
#to fix the switched GPL1/2 values
#assign1[assign1$value==1,] <- 3
#assign1[assign1$value==2,] <- 1
#assign1[assign1$value==3,] <- 2


#write.csv(cbind(dapc1$posterior, assign1), file="serdp_and_waddle_dapc_posterior_2clust_99cut_2PC_samptrim_90.csv")

vcf.meta <- cbind(vcf.meta, assign=assign1$value)

```

Now we use the assignments to make a PCA

```{r}

pca <- glPca(gl.Bd, nf = 3)
barplot(100*pca$eig/sum(pca$eig), col = heat.colors(50), main="PCA Eigenvalues")
title(ylab="Percent of variance\nexplained", line = 2)
title(xlab="Eigenvalues", line = 1)

pca.scores <- as.data.frame(pca$scores)

cols <- brewer.pal(n_distinct(assign1$value), "Set1")


p <- ggplot(pca.scores, aes(x=PC2, y=PC1, colour=as.factor(assign1$value), shape=Bd.meta$GROUP)) + 
  geom_point(size=5, alpha = 0.6 ) + 
  scale_color_manual(values = c("dark grey",cols[2],cols[1])) + 
  scale_shape_manual(values = c(15, 16, 17)) +
  stat_ellipse(aes(x=PC2, y=PC1, group=as.factor(assign1$value)),level = 0.95, size = 1) + 
  #geom_label_repel(aes(label = vcf.meta$New_Sample_ID), size = 3)+
  
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()

#PCA from figure 1
p

#to plot PCA with color based on missing data proportion
p_miss <- ggplot(pca.scores, aes(x=PC2, y=PC1, shape=Bd.meta$GROUP)) + 
  geom_point(aes(color = Bd.meta$Missing_data_snps), size = 5) +
  scale_shape_manual(values = c(15, 16, 17)) +
  scale_color_gradient(low = "yellow", high = "darkblue") +
  stat_ellipse(aes(x=PC2, y=PC1, group=as.factor(assign1$value)),level = 0.95, size = 1) +
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()

#PCA from figure S1
p_miss

```

Now we do the same but for each separate sample group


```{r}
#PA
gl.pa <- popsub(gl.Bd, sublist="Pennsylvania")

pca_pa <- glPca(gl.pa, nf = 3)
barplot(100*pca_pa$eig/sum(pca_pa$eig), col = heat.colors(50), main="PCA Eigenvalues")
title(ylab="Percent of variance\nexplained", line = 2)
title(xlab="Eigenvalues", line = 1)

pca.scores.pa <- as.data.frame(pca_pa$scores)

pa.vcf.meta<- filter(vcf.meta, GROUP=="Pennsylvania")

#for color by lineage and shape by species 
p_pa <- ggplot(pca.scores.pa, aes(x=PC2, y=PC1, colour=as.factor(pa.vcf.meta$assign), shape=pa.vcf.meta$Species_code)) + 
  geom_point(size=5, alpha = 0.9 ) + 
  scale_color_manual(values = c("dark grey",cols[2],cols[1])) + 
  scale_shape_manual(values = c(0,2,10,16,3,4,6,7,8,9,15,11,12)) +
  stat_ellipse(aes(x=PC2, y=PC1, group=as.factor(pa.vcf.meta$assign)),level = 0.95, size = 1, alpha = 0.4) + 
  #geom_label_repel(aes(label = vcf.meta$New_Sample_ID), size = 3)+
  
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()

p_pa


#for color by site and shape by species
cols2 <- brewer.pal(n_distinct(pa.vcf.meta$Site_code), "Set1")

p_pa2 <- ggplot(pca.scores.pa, aes(x=PC2, y=PC1, colour=as.factor(pa.vcf.meta$Site_code), shape=pa.vcf.meta$Species_code)) + 
  geom_point(size=5, alpha = 0.9 ) + 
  scale_color_manual(values = cols2) + 
  scale_shape_manual(values = c(0,2,10,16,3,4,6,7,8,9,15,11,12)) +
  stat_ellipse(aes(x=PC2, y=PC1, group=as.factor(pa.vcf.meta$Site_code)),level = 0.95, size = 1, alpha = 0.4) + 
  #geom_label_repel(aes(label = vcf.meta$New_Sample_ID), size = 3)+
  
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()

#PCA from Figure 3
p_pa2

```

Now let's make the PCA for the reference samples


```{r}
gl.ref <- popsub(gl.Bd, sublist="Reference")

pca_ref <- glPca(gl.ref, nf = 3)
barplot(100*pca_ref$eig/sum(pca_ref$eig), col = heat.colors(50), main="PCA Eigenvalues")
title(ylab="Percent of variance\nexplained", line = 2)
title(xlab="Eigenvalues", line = 1)

pca.scores.ref <- as.data.frame(pca_ref$scores)

ref.vcf.meta<- filter(vcf.meta, GROUP=="Reference")

p_ref1 <- ggplot(pca.scores.ref, aes(x=PC2, y=PC1, colour=as.factor(ref.vcf.meta$GENOASSIGN))) + 
  geom_point(size=5, alpha = 0.9 ) + 
  scale_color_manual(values = c(cols[2],cols[1])) + 
  scale_shape_manual(values = c(16,5,14,1,13)) +
  stat_ellipse(aes(x=PC2, y=PC1, group=as.factor(ref.vcf.meta$GENOASSIGN)),level = 0.95, size = 1, alpha = 0.4) + 
  geom_label_repel(aes(label = ref.vcf.meta$Sample_ID), size = 3)+
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()

p_ref1

```

Now let's make the PCA for Nevada


```{r}
#NV
gl.nv <- popsub(gl.Bd, sublist="Nevada")


pca_nv <- glPca(gl.nv, nf = 3)
barplot(100*pca_nv$eig/sum(pca_nv$eig), col = heat.colors(50), main="PCA Eigenvalues")
title(ylab="Percent of variance\nexplained", line = 2)
title(xlab="Eigenvalues", line = 1)

pca.scores.nv <- as.data.frame(pca_nv$scores)

nv.vcf.meta<- filter(vcf.meta, GROUP=="Nevada")

p_nv1 <- ggplot(pca.scores.nv, aes(x=PC2, y=PC1, colour=as.factor(nv.vcf.meta$assign), shape=nv.vcf.meta$species)) + 
  geom_point(size=5, alpha = 0.9 ) + 
  scale_color_manual(values = c(cols[2],cols[1])) + 
  scale_shape_manual(values = c(16,5,14,1,13)) +
  stat_ellipse(aes(x=PC2, y=PC1, group=as.factor(nv.vcf.meta$assign)),level = 0.95, size = 1, alpha = 0.4) + 
  #geom_label_repel(aes(label = vcf.meta$New_Sample_ID), size = 3)+
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()

p_nv1

#for color by site and shape by species
cols3 <- brewer.pal(n_distinct(nv.vcf.meta$Site), "Dark2")

p_nv2 <- ggplot(pca.scores.nv, aes(x=PC2, y=PC1, colour=as.factor(nv.vcf.meta$Site), shape=nv.vcf.meta$species)) + 
  geom_point(size=5, alpha = 0.9 ) + 
  scale_color_manual(values = cols3) + 
  scale_shape_manual(values = c(16,5,14,1,13)) +
  stat_ellipse(aes(x=PC2, y=PC1, group=as.factor(nv.vcf.meta$Site)),level = 0.95, size = 1, alpha = 0.4) + 
  #geom_label_repel(aes(label = vcf.meta$New_Sample_ID), size = 3)+
  geom_hline(yintercept = 0) + 
  geom_vline(xintercept = 0) + 
  theme_bw()

#pca from figure 2
p_nv2



```

Explore the "unknown" values. Is it because of missing data or is it because of some sort of mix or hybrid sample?


```{r}
#density plot from figure S1C
ggplot(data=vcf.meta, aes(x = Missing_data_snps, fill = as.factor(assign))) + geom_density(alpha = 0.5) +theme_bw()

```

Now we do AMOVA to test variation of genetic data based on site or species. 
Sourced from this [tutorial](https://grunwaldlab.github.io/poppr/reference/poppr.amova.html)

```{r}
#use the package poppr
my_genind <- vcfR2genind(Bd.VCF)

pop(my_genind) <- vcf.meta$GROUP
pa_genind <- popsub(my_genind, sublist="Pennsylvania")
nv_genind <- popsub(my_genind, sublist="Nevada")

pa.vcf.meta<- filter(vcf.meta, GROUP=="Pennsylvania")
nv.vcf.meta<- filter(vcf.meta, GROUP=="Nevada")

#set strata as site and species
pa_strata <- data.frame(cbind(pa.vcf.meta$Site_code,pa.vcf.meta$Species_code, paste(pa.vcf.meta$Site_code,pa.vcf.meta$Species_code, sep="_")))
colnames(pa_strata) <- c("Site","Species","Site_Species")

strata(pa_genind) <- pa_strata
pa_genclone <- as.genclone(pa_genind)

table(strata(pa_genclone, ~Site/Species, combine = FALSE))

PA_amova <- poppr.amova(pa_genclone, ~Site/Species)

PA_amovacc <- poppr.amova(pa_genclone, ~Site/Species, clonecorrect = TRUE)

write.table(PA_amova$componentsofcovariance, sep = ",", file = "PA_AMOVA.csv")


set.seed(1989)

PA_signif   <- randtest(PA_amova, nrepet = 999)
plot(PA_signif)

#PA_ccsignif <- randtest(PA_amovacc, nrepet = 999)
#plot(PA_ccsignif)

#NOW FOR NV

#set strata as site and species
nv_strata <- data.frame(cbind(nv.vcf.meta$Site_code,nv.vcf.meta$species, paste(nv.vcf.meta$Site_code,nv.vcf.meta$species, sep="_")))
colnames(nv_strata) <- c("Site","Species","Site_Species")

strata(nv_genind) <- nv_strata
nv_genclone <- as.genclone(nv_genind)

table(strata(nv_genclone, ~Site/Species, combine = FALSE))

NV_amova <- poppr.amova(nv_genclone, ~Site/Species)

NV_amovacc <- poppr.amova(nv_genclone, ~Site/Species, clonecorrect = TRUE)

set.seed(1989)

NV_signif   <- randtest(NV_amova, nrepet = 999)
plot(NV_signif)

#NV_ccsignif <- randtest(NV_amovacc, nrepet = 999)
#plot(NV_ccsignif)

write.table(NV_amova$statphi, sep = ",", file = "NV_AMOVA_stat.csv")

```

